Subsidiary Recursion in Coq

### 2022-02-06 paper/paper.tex
applying some of Ben's edits by hand

%TODO mandatory: add short abstract of the document
\begin{abstract}
  This paper describes a functor-generic derivation in Coq of
  subsidiary recursion.  On this recursion scheme, inner recursions
  may be initiated within outer ones, in such a way that outer
  recursive calls may be made on the results from inner ones.  The
  derivation utilizes a novel (necessarily weakened) form of
  positive-recursive types in Coq, dubbed retractive-positive
  recursive types.  A corresponding form of induction is also
  supported.  The method is demonstrated through several examples.
\end{abstract}

@@

Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
Lean and others are terminating recursive functions over user-declared
inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
usually enforced by a syntactic check for structural decrease, which
is sufficient for many basic functions.  For example, the
\texttt{span} function from Haskell's prelude (\verb|Data.List|) takes
a list and returns a pair of the maximal prefix whose elements satisfy
a given predicate \verb|p|, and the remaining suffix:
\begin{minted}{haskell}
span :: (a -> Bool) -> [a] -> ([a],[a])
span _ []    =  ([], [])
span p (x:xs) = if p x
                then let (ys,zs) = span p xs in (x:ys,zs)
                else ([],x:xs)
\end{minted}
\noindent The sole recursive call is \verb|span p xs|. This call is
structurally decreasing, since it occurs in a clause where the input
list is of the form \verb|x:xs|.  In the appropriate syntax, this
definition can be accepted without additional effort by all the
previously mentioned provers.

This paper presents a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), while still preserving
the possibility of further invocations of the outer recursive
function.  As a simple example, consider the function \verb|wordsBy|
(\verb|Data.List.Extra|), which breaks a list into a set of its
maximal sublists whose elements do not satisfy a predicate \verb|p|.
For example, \verb|wordsBy isSpace " good day "| returns
\verb|["good","day"]|. The code for this function is in
Figure~\ref{fig:wordsBy}, where \verb|break p| is defined as
\verb|span (not . p)|. The first recursive call, \verb|wordsBy p tl|,
is structurally decreasing. In contrast, the second invocation of
\verb|wordsBy p| uses a value obtained from a call to \verb|span|.
This recursive call is not allowed under the structural termination
condition, but is permitted by subsidiary recursion.

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We emphasize that this is
a derivation within the type theory of Coq, and requires no axioms or
other modifications to Coq, except the \verb|-impredicative-set|
flag. Using this derivation, we present several example functions in
the vein of \verb|wordsBy|, and prove theorems about their behavior.  A nice
example is a definition of run-length encoding using \verb|span| as a
subsidiary recursion, where we prove that encoding and then decoding
returns the original list.  Our approach applies to the standard
datatypes in the Coq library, and does not require switching libraries
or datatype definitions.

An important technical novelty of our development is a derivation of a
weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
restrict datatypes $D$ to be strictly positive: in the argument types
of constructors of $D$, $D$ cannot occur to the left of any arrows.
Our derivation needs to use positive-recursive types, where $D$ may
also occur to the left of an even number (only) of arrows.  We present
a way to derive a weakened form of positive-recursive type that is
sufficient for our examples (Section~\ref{sec:mu}).  The weakening is
to require only that $F\ (\mu F)$ is a retract of $\mu F$.  (Usually
these types are isomorphic.)  Hence, we dub these
\textbf{retractive-positive} recursive types.  This weakening leads to
noncanonical elements of $\mu$, but we will see how to work around
this.  Our definition of retractive-positive recursive types makes
essential use of impredicative quantification, and hence is not legal
in predicative theories like Agda's.

We begin by summarizing the interface our derivation provides for
subsidiary recursion (Section~\ref{sec:interface}), and we then
present some examples (Section~\ref{sec:examples}).  We next explain
the implementation of our interface (Section~\ref{sec:deriv}),
including our definition of retractive-positive recursive types
% (Section~\ref{sec:mu})
.  We then cover the interface for subsidiary
induction (Section~\ref{sec:interfacei}), and give some example proofs
that use it (Section~\ref{sec:examplesi}). We conclude with a discussion
of related work (Section~\ref{sec:related}).

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles (cf.~\cite{traytel12,cockett92,hagino87}). Under
this approach, inductive datatypes are equipped with a signature
functor which explains how to 'unroll' one layer of the type. The
signature functor for lists is:
\begin{minted}{coq}
Inductive ListF (X : Set) : Set :=
| Nil : ListF X
| Cons : A -> X -> ListF X.
\end{minted}
Transformations on data with functor $F$ are described as
\emph{algebras} which are then \textit{folded} over data.  The
simplest form of algebras are morphisms from $F\ A$ to $A$, for some
carrier object $A$.  From a programming perspective, an $F$-algebra is
given input of type $F\ A$, and must compute a result of type $A$. The
$ListF$-algebra for computing the length of a list is:
\begin{minted}{coq}
Definition LengthAlg (l : ListF nat) : nat :=
match l with
| Nil => 0
| Cons a len' => 1 + len'
end.
\end{minted}
Algebras for our subsidiary recursion are more complex than
$F$-algebras.  Let us begin with an informal explanation.  For reasons
we will explain further below, the carrier of the algebra will be a
functor \verb|X : Set -> Set|.  Our algebras are presented with:

@@

We use Coq inductive types for the signature functors \verb|F| of
various datatypes. This allows algebras to use Coq's
pattern-matching on the subdata structure \verb|d|, enabling a style
of coding that retains a similar feel to structural recursion. Unlike
structurally recursive functions, however, this interface uses a
type-based (and hence compositional) criteria to ensure termination.

We have previously dubbed this interface a \emph{recursion
  universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have a kind of space (here, \verb|R|), which one
cannot escape using the provided operations.  Other examples include the
ordinal $\epsilon_0$ and operator $\omega^-$, and the physical
universe and traveling at the speed of light.  Staying in the
recursion universe is good, because we may recurse (via \verb|rec|) on
any value of type \verb|R|.  Some points must still be explained: in
particular why \verb|X| has type \verb|Set -> Set|, and the definition
of \verb|FoldT|. We consider those details next.

@@

\begin{figure}
\begin{minted}{coq}
Definition List := Subrec ListF.
Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{minted}
  \caption{Some basics from \texttt{List.v}, specializing the
    functor-generic derivation of subsidiary recursion to
    lists parameterized by an element type \texttt{A}  (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

The type over which one can recurse using our subsidiary
recursion scheme is called \verb|Subrec|.  It is parametrized by a signature
functor \verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
\verb|inn : F Subrec -> Subrec|, which behaves computationally like a
constructor.  We will later derive an induction principle for this type
(Section~\ref{sec:interfacei}). The definition of \verb|Subrec| uses
retractive-positive recursive types, to take a fixed-point of a
construction based on \verb|F|.  We present these recursive types in
Section~\ref{sec:mu} below.

Our examples specialize this scheme to the case of polymorphic lists,
presented in Figure~\ref{fig:listf}.  In general, to use our
development to get subsidiary recursion over some datatype, one must
define a signature functor for the datatype, e.g., \verb|ListF| from
the start of this section. We define \verb|List| to be
\verb|Subrec| with \verb|F| instantiated to \verb|ListF A|.  This
type \verb|List| is not to be confused with the type \verb|list| of
lists from Coq's standard library.  As noted previously, our development
is meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and typings for conversion functions
between \verb|List| and \verb|list| (we elide their definitions).

@@

We need a fixed-point here due to the \verb|fold| parameter
\verb|AlgF| expecting an argument of type \verb|FoldT Alg R|.  This is the type for fold functions which
apply algebras (\verb|Alg|) to data of type \verb|R|.  The variable \verb|Alg| occurs only positively (but not
strictly positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

@@

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
nothing about \verb|R| except that it supports the
\verb|fold| and \verb|rec| operations that follow. The first is a local
\verb|fold| function, which allows us to fold another algebra over data
of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
recursions.  The second, \verb|rec|, is used to make recursive calls
on values of type \verb|R|.

As noted already, for subsidiary recursion, algebras have a carrier
\verb|X| which depends (functorially) on a type.  When we fold an
algebra using a fold function (either global or local) of type
\verb|FoldT Alg C|, (i) recursive calls may compute a result of type
\verb|X R|, mentioning the abstract type \verb|R| inside the recursion;
and (ii) outside that recursion, the result will have type \verb|X C|.
Having a functor for the carrier of the algebra gives us the
flexibility to type results inside a recursion with the abstract type
\verb|R|, but view those results as having the type \verb|C| outside
the recursion.  The function \verb|fold| in Figure~\ref{fig:algf} initiates
top-level folds.  We can also have functions
between \verb|Alg| and its \verb|AlgF|-unfolding.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

@@

This first example does not invoke subsidiary recursions, but will
itself be used as a subsidiary recursion in other examples to follow.
Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix.  This is the typing for a top-level recursion.  More
generally, though, given a type \verb|R : Set| along with a
fold function for that type (i.e., of type
\verb|FoldT (Alg (ListF A)) R|), we would like to map an input list of
type \verb|R| to a pair of type \verb|list A * R|.  The first
component of this pair is going to be built up from scratch, and so
cannot have type \verb|R|; we cannot statically ensure that outer
recursions on it are legal.  But the second component will be a
subdatum of the input list, and so can still have type \verb|R|,
enabling outer recursive calls.  So we want:
\begin{minted}{coq}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{minted}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{minted}{coq}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{minted}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|ListF A R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{minted}{coq}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{minted}
\noindent The idea is that the recursion will signal if it is in the
one tricky case where \verb|p| does not match the first element, by
returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This suffix is available to the algebra
in the subdata structure of type \verb|ListF A R|.

@@

Figure~\ref{fig:spanalg} show the algebra
\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
\verb|SpanF| of the required type \verb|Set -> Set|.  We use
\verb|rollAlg| to create an algebra from something whose type is an
application of \verb|AlgF|.  This takes in all the components of the
recursion universe: the abstract type \verb|R|, the fold function
(\verb|fo|) for any subsidiary recursions (not needed here), a
function we choose to name \verb|span| for making recursive calls, and
finally \verb|xs : ListF A R|.  The algebra pattern-matches on this
\verb|xs|.  In the cases where it is empty or where its head
(\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
This signals to the caller that we really wished to return
\verb|([],xs)|, but could not because we do not have \verb|xs| at type
\verb|R|.  If the head does satisfy \verb|p|, then we recurse on the
tail (\verb|tl : R|) by calling the provided
\verb|span : R -> SpanF R|.  If \verb|span tl| returns
\verb|SpanNoMatch|, that means that we should make \verb|tl| the
suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
have \verb|tl : R| here, so we can do this.  In either case (for
return value of \verb|span tl|), we add the head to the front of the
prefix.

@@

Let us now see how to write \verb|wordsBy|, our example function from
Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.  The code
is in Figure~\ref{fig:wordsby}, assuming a type \verb|A : Set|.  The
setup is similar to that for \verb|span|.  We first define an algebra
\verb|WordsBy|, parametrized by the predicate \verb|p|, of type
\verb|Alg (ListF A) (Const (list (list A)))|.  This says that
\verb|WordsBy p| is an algebra (\verb|Alg|) for the \verb|ListF A|
functor, with carrier \verb|Const (list (list A))|.  \verb|Const| is
a combinator for creating the object part of constant functors;
\verb|FunConst| creates the morphism part (i.e., the \verb|fmap|
function).  We use \verb|Const| where the return type of the algebra
will not depend on its abstract type \verb|R|.  Here, we are
constructing from scratch a list of lists, so it will not be legal to
recurse on the list itself, or its (list) elements.  So we just use
the \verb|list| type of Coq's standard library.

The code for \verb|WordsBy| is essentially
the same as what we saw in Section~\ref{sec:intro}.  We pattern match
on \verb|xs : ListF A R|.  Recall that for this function, we are
trying to drop elements which satisfy \verb|p|, and return a list of
the sublists between maximal sequences of such elements.  In the
\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we are
supposed to drop it and recurse.  This is legal, because \verb|tl : R|
and \verb|wordsBy : R -> list (list A)|.  In the \verb|else| case,
we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
does not satisfy \verb|p|, and the remaining suffix \verb|z|.

Here we see the benefit of our approach.  From Figure~\ref{fig:span},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
comes from the type \verb|FoldT (ListF A) Alg R| of
\verb|fo|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
can indeed apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.  The figure also shows the code for the subsidiary
recursion \verb|wordsByr|.

\begin{figure}
\begin{minted}{coq}
  Definition WordsBy(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs =>
      match xs with
        Nil => []
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p) xs.
\end{minted}
\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
  fold (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

In this example, we see how to write a combinator that factors out a
subsidiary recursion.  The Haskell library \verb|Data.List.Extra| has
a function \verb|repeatedly|, defined essentially as in
Figure~\ref{fig:mapthroughhs}, though we attempt a more informative
name.  This is like the standard \verb|map| function on lists, except
that the function \verb|f| that we are mapping (or ``mapping
through'') takes in not just the current element \verb|a|, but also
the tail \verb|as|.  It then returns the value \verb|b| to include in
the output list, and whatever other list it wishes, upon which
\verb|mapThrough| will recurse.

>>>>

%TODO mandatory: add short abstract of the document
\begin{abstract}
  This paper describes a functor-generic derivation in Coq of
  subsidiary recursion.  With this recursion scheme, inner recursions
  may be initiated within outer ones, in such a way that outer
  recursive calls may be made on results from inner ones.  The
  derivation utilizes a novel (necessarily weakened) form of
  positive-recursive types in Coq, dubbed retractive-positive
  recursive types.  A corresponding form of induction is also
  supported.  The method is demonstrated through several examples.
\end{abstract}

@@

Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
Lean and others are terminating recursive functions over user-declared
inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
usually enforced by a syntactic check for structural decrease, which
is sufficient for many basic functions.  For example, the
\texttt{span} function from Haskell's prelude (\verb|Data.List|) takes
a list and returns a pair of the maximal prefix whose elements satisfy
a given predicate \verb|p|, and the remaining suffix:
\begin{minted}{haskell}
span :: (a -> Bool) -> [a] -> ([a],[a])
span _ []    =  ([], [])
span p (x:xs) = if p x
                then let (ys,zs) = span p xs in (x:ys,zs)
                else ([],x:xs)
\end{minted}
\noindent The sole recursive call is \verb|span p xs|, and it occurs
in a clause where the input list is of the form \verb|x:xs|.  Hence it
is structurally decreasing.  In the appropriate syntax, this
definition can be accepted without additional effort by all the
mentioned provers.

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy|
(\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
elements do not satisfy a predicate \verb|p|.  For example,
\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|.
Code is in Figure~\ref{fig:wordsBy}.  Recall that
\verb|break p| is equivalent to \verb|span (not . p)|. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination, but
will be permitted by subsidiary recursion.

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We emphasize that this is
a derivation within the type theory of Coq, and requires no axioms or
other modifications to Coq, except the \verb|-impredicative-set|
flag. Using this derivation, we present several example functions like
\verb|wordsBy|, and prove theorems about them.  A nice example is a
definition of run-length encoding using \verb|span| as a subsidiary
recursion, where we prove that encoding and then decoding returns the
original list.  Our approach applies to the standard datatypes in the
Coq library, and does not require switching libraries or datatype
definitions.

An important technical novelty is a derivation of a weakened form of
positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
datatypes $D$ to be strictly positive: in the input types of
constructors of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left of an even number (only) of arrows.  We present a way to
derive a weakened form of positive-recursive type that is sufficient
for our examples (Section~\ref{sec:mu}).  The weakening is to require
only that $F\ (\mu F)$ is a retract of $\mu F$.  Usually, these types are
isomorphic.  Hence, we dub these \textbf{retractive-positive}
recursive types.  This weakening leads to noncanonical elements of
$\mu$, but we will see how to work around this.  Our definition of
retractive-positive recursive types makes essential use of
impredicative quantification, and hence is not legal in predicative
theories like Agda's.

We begin by summarizing the interface our derivation provides for
subsidiary recursion (Section~\ref{sec:interface}), and then see
examples (Section~\ref{sec:examples}).  We next explain how the
interface is actually implemented (Section~\ref{sec:deriv}), including
our retractive-positive recursive types (Section~\ref{sec:mu}).  The
interface for subsidiary induction is covered next
(Section~\ref{sec:interfacei}), and example proofs using it
(Section~\ref{sec:examplesi}).  Related work is discussed in
Section~\ref{sec:related}.

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles (cf.~\cite{traytel12,cockett92,hagino87}).  With
this approach, one describes transformations to be performed on data
as \emph{algebras}, which can then be \textit{folded} over data.  The
simplest form of algebras, namely $F$-algebras for functor $F$ (called
the \emph{signature functor} of the datatype), are morphisms from
$F\ A$ to $A$, for carrier object $A$.  From a programming
perspective, an $F$-algebra is given input of type $F\ A$, and must
compute a result of type $A$.  An example of $F$ is the signature
functor for lists, which we will use below:

\begin{minted}{coq}
Inductive ListF(X : Set) : Set :=
| Nil : ListF X
| Cons : A -> X -> ListF X.
\end{minted}

Algebras for our subsidiary recursion are more complex than
$F$-algebras.  Let us begin with an informal explanation.  For reasons
we will explain further below, the carrier of the algebra will be a
functor \verb|X : Set -> Set|.  The algebra is presented with:

@@

We will use Coq inductive types for the signature functors \verb|F| of
various datatypes. This allows algebras to use Coq's
pattern-matching on the subdata structure \verb|d|.  So the style of
coding against this interface retains a similar feel to structural
recursion.  Unlike with structural termination, though, the interface
here is type-based and hence compositional.

We have previously dubbed this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have a kind of space (here, \verb|R|), which one
cannot escape using certain operations.  Other examples are the
ordinal $\epsilon_0$ and $\omega^-$, and the physical universe and
traveling at the speed of light.  Staying in the recursion universe is
good, because we may recurse (via \verb|rec|) on any value of type
\verb|R|.  Some points must still be explained: why \verb|X| has type
\verb|Set -> Set|, and the definition of \verb|FoldT|.  Let us see
these details next.

@@

\begin{figure}
\begin{minted}{coq}
Definition List := Subrec ListF.
Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{minted}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists parametrized by an element type \texttt{A} (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

The type over which one can recurse using our scheme of subsidiary recursion is
called \verb|Subrec|.  It is parametrized by a signature functor
\verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
\verb|inn : F Subrec -> Subrec|, which behaves computationally like a
constructor.  We will later derive an induction principle for this
type (Section~\ref{sec:interfacei}). The definition of \verb|Subrec|
uses retractive-positive recursive types, to take a fixed-point of a
construction based on \verb|F|.  We present these recursive types in
Section~\ref{sec:mu} below.

In this paper, we use just the specialization to the case of lists,
with signature functor \verb|ListF A| shown above.  The parameter
\verb|A| is the type for the list elements.  \verb|List| is then
defined to be \verb|Subrec|, with \verb|F| instantiated to
\verb|ListF A|. In general, to use our development to get subsidiary
recursion over some datatype, one must define a signature functor for
the datatype.  Note that \verb|List| is different from the type
\verb|list| of lists in Coq's standard library.  Our development is
meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and typings for conversion functions
between \verb|List| and \verb|list| (definitions elided).

@@

We need a fixed-point here because \verb|Alg| occurs in the definition
of \verb|Algf|.  This is an essential circularity, because we are
trying to express that algebras take in \verb|fold| functions, which
themselves may accept algebras.  The variable \verb|Alg| occurs
negatively in \verb|FoldT Alg R| which occurs negatively in
\verb|AlgF Alg X|.  Hence it occurs positively but not strictly
positively.  So we can indeed take a fixed-point of \verb|AlgF| to
define the constant \verb|Alg|.

@@

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
nothing about \verb|R| except that it supports the following
operations.  We have a local
\texttt{fold} function, which will allow us to fold another algebra over data
of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
recursions.  Then there is \verb|rec|, for recursive calls on data of
type \verb|R|.

As noted already, for subsidiary recursion, algebras have a carrier
\verb|X| which depends (functorially) on a type.  When we fold an
algebra using a fold function (either global or local) of type
\verb|FoldT Alg C|, (i) recursive calls may compute a result of type
\verb|X R|, mentioning the abstract type \verb|R| for that recursion;
and (ii) outside that recursion, the result will have type \verb|X C|.
Having a functor for the carrier of the algebra gives us the
flexibility to type results inside a recursion with the abstract type
\verb|R|, but view those results as having the type \verb|C| outside
the recursion.  The function \verb|fold| (Figure~\ref{fig:algf}) initiates
top-level folds.  We also can have functions 
between \verb|Alg| and its \verb|Algf|-unfolding.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

@@

This first example does not invoke subsidiary recursions, but will
itself be used as a subsidiary recursion in other examples to follow.
Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix.  This is the typing for a top-level recursion.  More
generally, though, given a type \verb|R : Set| along with a
fold function for that type (i.e., of type
\verb|FoldT (Alg (ListF A)) R|), we will map an input list of
type \verb|R| to a pair of type \verb|list A * R|.  The first
component of this pair is going to be built up from scratch, and so
cannot have type \verb|R|; we cannot statically ensure that outer
recursions on it are legal.  But the second component will always be a
subdatum of the input list, and so can still have type \verb|R|,
enabling outer recursive calls.  So we want:
\begin{minted}{coq}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{minted}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{minted}{coq}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{minted}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|ListF A R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{minted}{coq}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{minted}
\noindent The idea is that the recursion will return \verb|SpanNoMatch| to signal that it is in the
one tricky case where \verb|p| does not match the first element.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This suffix is available to the algebra
in the subdata structure of type \verb|ListF A R|.

@@

Figure~\ref{fig:spanalg} shows the algebra
\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
\verb|SpanF| of the required type \verb|Set -> Set|.  We use
\verb|rollAlg| to create an algebra from something whose type is an
application of \verb|AlgF|.  This takes in all the components of the
recursion universe: the abstract type \verb|R|, the fold function
(\verb|fo|) for any subsidiary recursions (not needed here), a
function we choose to name \verb|span| for making recursive calls, and
finally \verb|xs : ListF A R|.  The algebra pattern-matches on this
\verb|xs|.  In the cases where it is empty or where its head
(\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
This signals to the caller that we really wished to return
\verb|([],xs)|, but could not because we do not have \verb|xs| at type
\verb|R|.  If the head does satisfy \verb|p|, then we recurse on the
tail (\verb|tl : R|) by calling the provided
\verb|span : R -> SpanF R|.  If \verb|span tl| returns
\verb|SpanNoMatch|, that means that we should make \verb|tl| the
suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
have \verb|tl : R| here, so we can do this.  For either possible return
value of \verb|span tl|, we add the head to the front of the
prefix.

@@

We now consider how to write the \verb|wordsBy| function from
Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.  The code
is in Figure~\ref{fig:wordsby}, assuming a type \verb|A : Set|.  The
setup is similar to that for \verb|span|.  We first define an algebra
\verb|WordsByAlg| of type
\verb|Alg (ListF A) (Const (list (list A)))|, parametrized
by a predicate \verb|p|.  This type expresses that
\verb|WordsByAlg p| is an algebra (\verb|Alg|) for the \verb|ListF A|
functor, with carrier \verb|Const (list (list A))|.  \verb|Const| is
a combinator for creating the object part of constant functors;
\verb|FunConst| creates the morphism part (i.e., the \verb|fmap|
function).  We use \verb|Const| where the return type of the algebra
will not depend on its abstract type \verb|R|.  Since we are
constructing a list of lists from scratch, it will not be legal to
recurse on the list itself, or its (list) elements.  So we just use
the \verb|list| type of Coq's standard library.

The code for \verb|WordsByAlg| is essentially
the same as what we saw in Section~\ref{sec:intro}.  Recall that this function drops elements
that satisfy \verb|p|, and returns the list of
sublists between maximal sequences of such elements.  The algebra
pattern-matches on \verb|xs : ListF A R|. In the
\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we
drop it and recurse.  Legality of the recursive call follows by typing: \verb|tl : R|
has the type expected by \verb|wordsBy : R -> list (list A)|.  Otherwise,
we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
does not satisfy \verb|p|, and the remaining suffix \verb|z|.

Here we see the benefit of our approach.  From Figure~\ref{fig:span},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
comes from the type \verb|FoldT (ListF A) Alg R| of
\verb|fo|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  Thus, it is
legal to apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.  The figure also shows the code for the subsidiary
recursion \verb|wordsByr|.

\begin{figure}
\begin{minted}{coq}
  Definition WordsByAlg(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs => 
      match xs with
        Nil => [] 
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
\end{minted}
\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
  fold (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

This example shows how to write a combinator that factors out a
subsidiary recursion.  The Haskell library \verb|Data.List.Extra|
defines a function \verb|repeatedly| in essentially the same way as
\verb|mapThrough| in Figure~\ref{fig:mapthroughhs} (we propose a more
informative name).  This function behaves like the standard \verb|map|
function on lists, except that the function \verb|f| that we are
mapping (or ``mapping through'') takes in not just the current element
\verb|a|, but also the tail \verb|as|.  It then returns the value
\verb|b| to include in the output list, and whatever other list it
wishes, upon which \verb|mapThrough| will then recurse.

To write this combinator using our infrastructure for subsidiary
recursion, we need to supply the mapped function with the fold
function for \verb|mapThrough|'s recursion.  This is so that the
mapped function can initiate a subsidiary recursion, returning a value
in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  
The type we will use for mapped functions is:
\begin{minted}{coq}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{minted}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input.
